---
title: "Lab 1: Intro to R"
author: "Sam Reeves"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
```

### Exercise 1

```{r view-girls-counts}
arbuthnot$girls
```


### Exercise 2

It appears that the trend is positive!

```{r trend-girls}
ggplot(data = arbuthnot, aes(x = year, y = girls)) + geom_line()
```


### Exercise 3

It appears that the trend of boy-to-girl births is slightly negative!

```{r plot-prop-boys-arbuthnot}
arbuthnot <- arbuthnot %>%
  mutate(total = boys + girls) %>%
  mutate(ratio = boys / girls)

ggplot(data = arbuthnot, aes(x = year, y = ratio)) + geom_line()
```


### Exercise 4

years = 1940 - 2002
dimensions = 63 x 3
variables = "year"  "boys"  "girls"

```{r dim-present}
summarize(present, min = min(year), max = max(year))
dim(present)
colnames(present)
```


### Exercise 5

The totals are two orders of magnitude greater in the "present" dataset, but the ratio of boys to girls is very similar.

```{r count-compare}
present <- present %>%
  mutate(total = boys + girls) %>%
  mutate(ratio = boys / girls)

mean(arbuthnot$total)
mean(present$total)

mean(arbuthnot$ratio)
mean(present$ratio)
```


### Exercise 6

Yes, Arbuthnot's assertion that boys are born more often, does still hold up, though it tends towards even at the end of the dataset.

```{r plot-prop-boys-present}
ggplot(data = present, aes(x = year, y = ratio)) + geom_line()
```


### Exercise 7

1961 had the most total births in the set. There were 4,268,326

```{r find-max-total}
arrange(present, desc(total))
```

